Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-04 Manuscript/manuscript.Rmd
full edits

The distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries. Evidence suggests that regulatory elements and genes tend to interact more frequently within the same TAD [@symmons2014functional]. Likewise, emerging research has shown that particular binding sites are enriched at TAD boundaries. It has been established that the zinc finger protein, CTCF, is associated with TAD boundaries and is an architectural protein responsible for the organization of the genome within the nucleus [@ghirlando2016ctcf; @ong2014ctcf]. CTCF along with other transcription factors, including SMC1, SMC3, and RAD21 have been shown to colocalize together near TAD boundaries, leading to the recruitment of a cohesin ring [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Together, these components are hypothesized to be what causes loops in the DNA, contributing to higher order chromatin organization. Other binding proteins have been shown to mediate enhancer-promoter interactions, including the transcription factors YY1 and ZNF143, thereby influencing the delineation of topological domains [@weintraub2017yy1; @gonzalez2017znf143; @heidari2014genome]. Even more so, loci within a domain have been shown to interact more frequently with one another, exhibiting similar patterns of chromatin modifications, and long-range contact patterns [@rao20143d]. However, the growing number of annotated genomic features and the methods used to identify the ones assocaited with TAD boundaries still leaves much to be desired.

Current methods used to predict TAD boundaries rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries. However, this is a multi-faceted probelem, and many methods ignore key characteristics of domain data when building models and reporting performances. We aim to address these characteristics in order to create a more unified approach toward identifyig the genomic annotations most predictive of TAD boundaries.

A major ignored aspect of TAD boundary prediction is the introduction of heavily imbalanced classes resulting from the number of boundaries vs. nonboundary regions. In a classification setting, when one is attempting to discriminate between minority (regions with TAD boundaries) and majority classes (regions without TAD boundaries), the class imbalance issue will contribute to small imbalance ratios. The imbalance ratio is defined as the ratio of minority to majority classes [@Orriols-Puig2008]. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing].

In a study published in 2016 [@mourad2016computational], an L1-regularized multiple logistic regression model was proposed using transcription factor binding sites on GM12878 to discriminate between regions with and without TADs, and to identify the most predictive factors associated with boundaries. However, the class imbalance issue was not addressed. Even worse, when performing cross-validation, [@mourad2016computational] used the misclassification error rate of the model to establish the optimal L1-regularization lambda. They also reported ROC (receiver operating characteristic) curves and subsequent AUC (area under the curve) as measures of their model's performance. This is problematic in an imbalanced setting. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing]. For AUC, when the prevalence of an event is low--that is, a defined genomic region truly containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey; @wei2013role]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. Consequently, careful consideration must be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes [@saito2015precision]. Another threshold-specific measure to consider is a composite metric involving all quadrants of the confusion matirix, such as Matthews Correlation Coefficient (MCC) [@jeni2013facing; @chicco2017ten]. The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of positive 1 represents a perfect prediction, while a value of negative 1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.

In addition to employing suitable metrics for evaluating models validated on imbalanced data, there also exists several data level solutions. Data level solutions mostly constitute as re-sampling techniques, in which random sampling is used to either replicate or remove data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data]. It has been shown that quite simple re-sampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis].

Two studies published in 2015 and 2017 respectively [@huang2015predicting; @hong2017computational] used re-sampling techniques prior to modeling, and both implemented random under-sampling to balance classes. However, given the large disparity between the number of genomic regions labeled as containing a TAD boundary or not, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling, for instance random over-sampling or a weighted combination of both random under- and over-sampling together. Equally concerning is how the test set is treated in scenarios of re-sampling. It was found that [@huang2015predicting] performed re-sampling prior to splitting the data into training and testing. However, this is problematic because re-sampling prior to splitting can further introduce bias into the testing set, as it is no longer an accurate representation of real-world data [@wei2013role].

@@

Therefore, in this study, we aimed to develop an ensemble framework composed of different combinations of data driven characteristics used to address the issues presented above. We used our framework to systematically compare performances of random forest classifiers using a composite performance metric, MCC. Our goal was to determine a suitable combination of data re-sampling and feature engineering techniques that both corrected the inherent imbalanced classes and provided the best spatial representation linking genomic annotations to TAD boundaries. Additionally, we used our framework to directly compare the influence of Hi-C contact matrix resolution on model performances. We evaluated Hi-C data for both the GM12878 and K562 cell line obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolution.

For each resolution, we examined the inclusion of three different predictor types, including distance as the measure of association between genomic elements. We then tested four different re-sampling techniques aimed at creating balanced classes. This process was performed on chromosome-specific regions on the linear genome. Noninformative features were removed using a feature reduction algorithm. Lastly, random forest classifiers were built to assess model performance, and investigate which genomic annotations were most predictive of TAD boundaries for a given system from the complete ensemble.

@@

Biological replicate Hi-C data for the GM12878 and K562 cell lines were downloaded from GEO GSE63525 [@rao20143d] (Supplementary Table 1). The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. The start and end coordinates of all TADs were concatenated, sorted, and unique TAD boundary points were obtained for the autosomal chromosomes. TAD boundaries called on Chromosome 9 were omitted from all downstream analyses, for both cell lines, due to the sparcity of the contact matrix at the 10 kb resolution for the K562 cell line.

@@

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution for the corresponding contact matrix. As an example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. For the boundary data, in order to accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. That is, for 10 kb resolution data, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was then labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0). The process was repeated for each subsequent resolution.

@@

2. **Overlap Percent (OP)**: For each genomic bin, the percent of overlap between the width of the region assigned to a genomic annotation and the total width of the bin was calculated. We defined the feature width as the number of bases that overlapped with a particular genomic bin. The percent was then calculated by dividing the feature width by the bin width. If multiple overlaps existed, the feature width was defined as the sum of the total number of bases that overlapped with a particular bin. Note that the bin width remained constant for a given resolution (either 10 kb, 25 kb, 50 kb, or 100 kb). Values for bins with no overlaps were set to be 0.

@@

Architectural proteins tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To measure the pairwise correlation of each cell line specific annotation, we used the Jaccard Index. The Jaccard Index measures the ratio of the number of intersecting base pairs between two sets to the number of base pairs in the union, subtracted by the intersection [@favorov2012exploring]. The quantity ranges from 0.0 to 1.0, where 0.0 represents no overlap (no correlation) and 1.0 represent complete overlap (complete correlation).

In order to suitably reduce the feature space while also maintaining the inclusion of possibly colocalized annotations, we implemented an elastic-net regularization technique prior to modelling. The technique using a combination of the *l_{1}* and *l_{2}* penalties. For each elastic-net regularization, 10-fold cross-validation was used to tune the parameters that controlled the penalty terms. <!--The parameters that maximized the mean AUPRC across each cross-fold was chosen.--> All elastic-net algorithms were implemented in R using the *caret* package version 6.0 [@kuhn2012caret].

@@

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were built to compare model performances between re-sampling techniques and predictor types. <!--To reduce bias due to random dataset generation, 10-fold cross validation was used to tune the number of features to consider at each node and the number of ensemble trees to aggregate.--> The default number of features to consider at each node of the random forest algorithm was set as $\sqrt{p}$, where *p* refers to the number of predictors included in the model. Likewise, the number of ensemble trees to aggregate was set at 500. All random forest algorithms were implemented in R using the *caret* package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was validated on the test set.

@@

$$MCC = \frac{TP \times TN - FP \times FN}{\sqrt{ (TP + FP)(TP+FN)(TN+FP)(TN+FN) }}$$
<!--
$$Balanced \quad Accuracy = \dfrac{1}{2}\left( \dfrac{TP}{TP+FN} + \dfrac{TN}{TN+FP} \right)$$
-->

@@

To enable us to build a model that suitably discriminated between regions on the linear genome that were near TAD boundaries and regions that were not, we created a  resolution-specific response variable that indicated whether a genomic bin of unit width overlapped with a flanked TAD boundary ($Y=1$), or not ($Y=0$). The unit width of each bin on the linear genome corresponded to the resolution of the cell line specific contact matrix on which domains were called. Moreover, TAD boundaries were flanked on either side by the same unit length as the genomic bins to account for uncertainty in boundary location.

Expectedly, the median number of genomic bins, TADs called, and TAD boundaries increased with the increased resolution of cell-line specific Hi-C data (Table 1; Supplementary Tables 3). This correlation with resolution was found to be significant, as evidenced by large intraclass correllation coefficients (ICC) (Table 2; Supplementary Tables 4). However, the class imbalance problem, caused by the sparsity of TAD boundaries throughout the genome, was found to be much more consistant across resolutions in both cell lines. This imbalance was evidenced by low imbalance ratios, defined as the ratio of the number of genomic bins that overlapped with flanked TAD boundaries, to the much larger number of genomic bins that did not. This relationship was not significantly correlated with resoltuion for either cell line, indicating that there existed a severe disparity between classes at all resolutions of Hi-C data.

@@

To establish the feature space for all downstream models, we constructed 3 different predictor-types. These predictor-type specific annotations represented the data matrix, $X = \{X_1, X_2, ..., X_p\}$, where $p$ was the complete set of functional annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 2A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 2B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 2C). A $log_{2}}$ transformation was used as a normalization technique in order to account for the skewness present in the distance densities (Supplementary Figures 1-2).

@@

To systematically summarize and compare our results of data driven random forest models, we created an ensemble framework (Figure 3). Each system in the framework was assumed to be composed of a model constructed from a particular cell line, $L = \{GM12878, K562\}$. Within each cell line we evaluated 4 data resolutions, $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. We then considered 3 separate predictor types as the feature space including $P = \{OC, OP, Distance\}$. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9). Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.

@@

We then went on the evaluate the effect of class imbalance on model performance. In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 4A). We gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Here, the results were similar to that of OC types (Figure 4B). However, when including distance-type predictors, models were found to exhibit a marked improvement in MCC compared to the other two types (Figure 4C). Results were found to be consistently poor across resolutions and predictor types, when implemented on the K562 cell line (Supplementary Figure 3). Furthermore, we found that conventional metrics of performance (i.e. accuracy, sensitivity and specificity, and AUC) were less sensitive for models built on imbalanced data, regardless of cell line, resolution, or predictor-type (Supplementary Tables 6-7). [??? Need to know why PRCs give different results.]

@@

When assessing performances for models built on balanced data, we found that SMOTE re-sampling provided the optimal MCC consistently across both predictor type and resolution (Figure 4). Specifically, the most improvement in model performance was seen when combining SMOTE re-sampling with distance-type predictors. This was also the optimal combination for the K562 cell line (Supplementary Figure 3). These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that is built on 10 kb resolution data, with distance-type predictors, and incorporates both random under- and over-sampling. Moreover, there was a clear degredation in model performance as data resolution decreased. Likewise, performances exhibited much more variability across chromosomes when built on lower resolutions.

@@

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites), across resolution. We found that for random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the cell-line specific transcription factor binding sites, and was consistent across resolutions (Figure 5; Supplementary Figure 4).

@@

We then sought to determine which of the transcription factor binding sites were most influential in predicting TAD boundaries. We found that the cell line specific binding sites associated with SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were given consistently high predictive importance, across all resolutions (Figure 7; Supplementary Figure 6). The SMC3, CTCF, and RAD21 binding sites are known factors associated with the cohesin protein complex, which is hypothesized to be what creates loops in the DNA  [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likwise, ZNF143 and YY1 are known structural regulators of enhancer-promoter interactions that also contribute to looping of the linear genome [@weintraub2017yy1; @gonzalez2017znf143]. With the identification of these factors known to be enriched at TAD boundaries, we are able to confirm the validity of our model building process. Additionally, we were also able to identify some lesser known cell line specific transcription factors as predictive of TAD boundaries including the MYC-associated zing-finger protein (MAZ), jun-D proto-oncogene (JUND), and ELF1. These transcription factor binding sites tended to be given higher predictive importance at lower resolutions. In fact, at lower resolutions, the predictive importance was found to be much more evenly dispersed across the top transcription factor binding sites, than for higher resolutions. This is again suggestive of the fact that the spatial relationship between functional annotations, specifically transcription factor binding sites, and TAD boundaries changes at different Hi-C data resolutions.

>>>>

The distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries. Evidence suggests that regulatory elements and genes tend to interact more frequently within the same TAD [@symmons2014functional]. Likewise, emerging research has shown that particular binding sites are enriched at TAD boundaries. It has been established that the zinc finger protein, CTCF, is associated with TAD boundaries and is an architectural protein responsible for the organization of the genome within the nucleus [@ghirlando2016ctcf; @ong2014ctcf]. CTCF along with other transcription factors, including SMC1, SMC3, and RAD21 have been shown to colocalize together near TAD boundaries, leading to the recruitment of a cohesin ring [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Together, these components are hypothesized to be what causes loops in the DNA, contributing to higher order chromatin organization. Other binding proteins have been shown to mediate enhancer-promoter interactions, including the transcription factors YY1 and ZNF143, thereby influencing the delineation of topological domains [@weintraub2017yy1; @gonzalez2017znf143; @heidari2014genome]. Even more so, loci within a domain have been shown to interact more frequently with one another, exhibiting similar patterns of chromatin modifications, and long-range contact patterns [@rao20143d]. However, the growing number of annotated genomic features and the methods used to identify the ones associated with TAD boundaries still leaves much to be desired.

Current methods used to predict TAD boundaries rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries. However, this is a multi-faceted probelem, and many methods ignore key characteristics [?? which ones ??] of domain data when building models and reporting performances. We aim to address these characteristics in order to create a more unified approach toward identifying the genomic annotations most predictive of TAD boundaries.

A major ignored aspect of TAD boundary prediction is the introduction of heavily imbalanced classes resulting from the number of boundaries vs. nonboundary regions. In a classification setting, when one is attempting to discriminate between minority (regions with TAD boundaries) and majority classes (regions without TAD boundaries), the class imbalance issue will contribute to small imbalance ratios. The imbalance ratio is defined as the ratio of minority to majority classes [@Orriols-Puig2008]. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing] [?? This sentence is very difficult to understand. Don't see connection between the bias and the error rate ??].

In a study published in 2016 [@mourad2016computational], an L1-regularized multiple logistic regression model was proposed using transcription factor binding sites on GM12878 to discriminate between regions with and without TADs, and to identify the most predictive factors associated with boundaries. However, the class imbalance issue was not addressed. Even worse, when performing cross-validation, [@mourad2016computational] used the misclassification error rate of the model to establish the optimal L1-regularization lambda. They also reported ROC (receiver operating characteristic) curves and subsequent AUC (area under the curve) as measures of their model's performance. This is problematic in an imbalanced setting. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing] [?? This is the second mention of "biased towards the majority class to minimize the overall error rate" without strong explanation of what this means ??]. For AUC, when the prevalence of an event is low--that is, a defined genomic region truly containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey; @wei2013role]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance [?? How so? does it inflate it or deflate it or what ??]. Consequently, careful consideration must be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases [?? Need to identify what you mean by "total number of cases." Is it just true positive divided by total number of possible boundaries ??]. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes [@saito2015precision]. Another threshold-specific measure to consider is a composite metric involving all quadrants of the confusion matrix, such as Matthews Correlation Coefficient (MCC) [@jeni2013facing; @chicco2017ten] [?? MCC actually has all of the same components as AUC. AUC is a function of sensitivity and specificity which combined contain all quandrants ??]. The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.

We can account for imbalance at the data level in addition to at the evaluation level [?? I rewrote this sentence. Was very wordy and difficult to follow ??]. Data level solutions mostly constitute as [?? I'd change "constiute as", unorthodox word choice ??] re-sampling techniques, in which random sampling is used to either replicate or remove data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data]. It has been shown that quite simple re-sampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. [?? It feels like data level should come before evaluation level ??]

Two studies published in 2015 and 2017 respectively [@huang2015predicting; @hong2017computational] used re-sampling techniques prior to modeling, and both implemented random under-sampling to balance classes. However, given the large disparity between the number of genomic regions labeled as containing a TAD boundary or not, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Equally concerning is how the test set is treated in scenarios of re-sampling. It was found that [@huang2015predicting] performed re-sampling prior to splitting the data into training and testing. However, this is problematic because re-sampling prior to splitting can further introduce bias into the testing set, as it is no longer an accurate representation of real-world data [@wei2013role].

@@

In this study, we aimed to develop an ensemble framework composed of different combinations of data driven characteristics used to address the issues presented above. We used our framework to systematically compare performances of random forest classifiers using a composite performance metric, MCC. Our goal was to determine a suitable combination of data re-sampling and feature engineering techniques that both corrected the inherent imbalanced classes and provided the best spatial representation [?? Not sure what spatial representation means in this context ??] linking genomic annotations to TAD boundaries. Additionally, we used our framework to directly compare the influence of Hi-C contact matrix resolution on model performances. We evaluated Hi-C data for both the GM12878 and K562 cell line obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolution. [?? It feels like some stronger statements need to be here. I would make a statement about predicting TAD regions using genomic annotations instead of using terms like "linking". It feels like the focus should be on this and not model selection which occurs in any analysis ??]

For each resolution, we examined the inclusion of three different predictor types [?? Maybe list all 3 (.e.g "For each resolution, we examined the inclusion of three different predictor types: counts, percentages and distances.") ??], including distance as the measure of association between genomic elements. We then tested four different re-sampling techniques aimed at creating balanced classes. This process was performed on chromosome-specific regions on the linear genome. Noninformative features were removed using a feature reduction algorithm. Lastly, random forest classifiers were built to assess model performance, and investigate which genomic annotations were most predictive of TAD boundaries for a given system from the complete ensemble.

@@

Biological replicate Hi-C data for the GM12878 and K562 cell lines were downloaded from GEO GSE63525 [@rao20143d] (Supplementary Table 1). The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. The start and end coordinates of all TADs were concatenated, sorted, and unique TAD boundary points were obtained for the autosomal chromosomes [?? Don't think sentence is necessary ??]. TAD boundaries called on chromosome 9 were omitted from all downstream analyses, for both cell lines, due to the sparsity of the contact matrix at the 10 kb resolution for the K562 cell line.

@@

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution of the corresponding contact matrix. As an example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. For the boundary data, in order to accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. That is, for 10 kb resolution data, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was then labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0). The process was repeated for each subsequent resolution [?? Don't think this sentence is needed ??].

@@

2. **Overlap Percent (OP)**: For each genomic bin, the percent of overlap between the width of the region assigned to a genomic annotation and the total width of the bin was calculated. We defined the feature width as the number of bases that overlapped with a particular genomic bin. The percent was then calculated by dividing the feature width by the bin width. If multiple overlaps existed, the feature width was defined as the sum of the total number of bases that overlapped with a particular bin. Note that the bin width remained constant for a given resolution (either 10 kb, 25 kb, 50 kb, or 100 kb). Values for bins with no overlaps were set to be 0. [?? This entire section feels extremely wordy. Probably can reduce to a couple sentences ??]

@@

Architectural proteins tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To measure the pairwise correlation of each cell line specific annotation, we used the Jaccard Index. The Jaccard Index measures the ratio of the number of intersecting base pairs between two sets to the number of base pairs in the union, subtracted by the intersection [@favorov2012exploring]. The quantity ranges from 0.0 to 1.0, where 0.0 represents no overlap (no correlation) and 1.0 represent complete overlap (complete correlation). [?? Jaccard is a distance measure, I would avoid using the term correlation here ??]

In order to suitably reduce the feature space while also maintaining the inclusion of possibly colocalized annotations, we implemented an elastic-net regularization technique prior to modelling. The technique using a combination of the *l_{1}* and *l_{2}* penalties. For each elastic-net regularization [?? It would help to describe why you used elastic-net and not ridge regression or lasso. Are you particularly concerned about bias/variance ??], 10-fold cross-validation was used to tune the parameters that controlled the penalty terms. <!--The parameters that maximized the mean AUPRC across each cross-fold was chosen.--> All elastic-net algorithms were implemented in R using the *caret* package version 6.0 [@kuhn2012caret].

@@

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were built to compare model performances between re-sampling techniques and predictor types. <!--To reduce bias due to random dataset generation, 10-fold cross validation was used to tune the number of features to consider at each node and the number of ensemble trees to aggregate.--> The default number of features to consider at each node of the random forest algorithm was set as $\sqrt{p}$, where *p* refers to the number of predictors included in the model. Likewise, the number of ensemble trees to aggregate was set at 500. All random forest algorithms were implemented in R using the *caret* package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was validated on the test set. [?? Don't think this sentence is necessary. Its implied you will run the model ??]

@@

$$
MCC = \frac{TP \times TN - FP \times FN}{\sqrt{ (TP + FP)(TP+FN)(TN+FP)(TN+FN) }}
$$
<!--
$$Balanced \quad Accuracy = \dfrac{1}{2}\left( \dfrac{TP}{TP+FN} + \dfrac{TN}{TN+FP} \right)$$
-->

@@

To enable us to build a model that suitably discriminated between regions on the linear genome that were near TAD boundaries and regions that were not, we created a  resolution-specific response variable that indicated whether a genomic bin of unit width overlapped with a flanked TAD boundary ($Y=1$), or not ($Y=0$) [?? extremely wordy sentence, can reduce ??]. The unit width of each bin on the linear genome corresponded to the resolution of the contact matrix on which domains were called. Moreover, TAD boundaries were flanked on either side by the same unit length as the genomic bins to account for uncertainty in boundary location.

Expectedly, the median number of genomic bins, TADs called, and TAD boundaries increased with the increased resolution of cell-line specific Hi-C data (Table 1; Supplementary Tables 3). This correlation [?? What correlation ??] with resolution was found to be significant, as evidenced by large intraclass correlation coefficients (ICC) (Table 2; Supplementary Tables 4). However, the class imbalance problem, caused by the sparsity of TAD boundaries throughout the genome, was found to be much more consistant across resolutions in both cell lines. This imbalance was evidenced by low imbalance ratios, defined as the ratio of the number of genomic bins that overlapped with flanked TAD boundaries, to the much larger number of genomic bins that did not. This relationship was not significantly correlated with resoltuion for either cell line, indicating that there existed a severe disparity between classes at all resolutions of Hi-C data.

@@

To establish the feature space for all downstream models, we constructed 3 different predictor-types. These predictor-type specific annotations represented the data matrix, $X = \{X_1, X_2, ..., X_p\}$, where $p$ was the complete set of functional annotations [?? $p$ is actually the total number of functional annotations, $X_i$ is the functional annotation itself ??]. Overlap counts was the least granular [?? Maybe define granular ??] predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 2A). Overlap percents was a more refined [?? what is refined in this context ??] predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 2B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 2C). A $log_{2}}$ transformation was used as a normalization technique in order to account for the skewness present in the distance densities (Supplementary Figures 1-2). [?? Why would skewness matter, why not use raw data ??]

@@

To systematically summarize and compare our results of data driven random forest models, we created an ensemble framework (Figure 3). Each system in the framework was assumed to be composed of a model constructed from a particular cell line, $L = \{GM12878, K562\}$. Within each cell line we evaluated 4 data resolutions, $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. We then considered 3 separate predictor types as the feature space including $P = \{OC, OP, Distance\}$. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9). Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials. [?? Where does variable selection come into play here??]

@@

We then went on the evaluate the effect of class imbalance on model performance. In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 4A). We gradually increased the granularity [?? Really confused by the constant use of granularity. Why is it needed ??] by assessing models built on OP predictor types, while still not performing any data resampling. Here, the results were similar to that of OC types (Figure 4B). However, when including distance-type predictors, models were found to exhibit a marked improvement in MCC compared to the other two types (Figure 4C). Results were found to be consistently poor across resolutions and predictor types, when implemented on the K562 cell line (Supplementary Figure 3). Furthermore, we found that conventional metrics of performance (i.e. accuracy, sensitivity and specificity, and AUC) were less sensitive for models built on imbalanced data, regardless of cell line, resolution, or predictor-type (Supplementary Tables 6-7). [??? Need to know why PRCs give different results.]

@@

When assessing performances for models built on balanced data, we found that SMOTE re-sampling provided the optimal MCC consistently across both predictor type and resolution (Figure 4). Specifically, the most improvement in model performance [?? Over what ??] was seen when combining SMOTE re-sampling with distance-type predictors. This was also the optimal combination for the K562 cell line (Supplementary Figure 3). These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries [?? Is this what you are doing? Or are you identifying TAD boundaries themselves ??], is one that is built on 10 kb resolution data, with distance-type predictors, and incorporates both random under- and over-sampling. Moreover, there was a clear degredation in model performance as data resolution decreased. Likewise, performances exhibited much more variability across chromosomes when built on lower resolutions.

@@

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models [?? In methods you say you do "feature selection" and then test model types. This implies the opposite ??]. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites), across resolution. We found that for random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the cell-line specific transcription factor binding sites, and was consistent across resolutions (Figure 5; Supplementary Figure 4).

@@

We then sought to determine which of the transcription factor binding sites were most influential in predicting TAD boundaries. We found that the cell line specific binding sites associated with SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were given consistently high predictive importance, across all resolutions [?? Is this after elastic net? Elastic net distributes coefficients between correlated variables and it may affect variable importance ??] (Figure 7; Supplementary Figure 6). The SMC3, CTCF, and RAD21 binding sites are known factors associated with the cohesin protein complex, which is hypothesized to be what creates loops in the DNA  [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, ZNF143 and YY1 are known structural regulators of enhancer-promoter interactions that also contribute to looping of the linear genome [@weintraub2017yy1; @gonzalez2017znf143]. With the identification of these factors known to be enriched at TAD boundaries, we are able to confirm the validity of our model building process. Additionally, we were also able to identify some lesser known cell line specific transcription factors as predictive of TAD boundaries including the MYC-associated zing-finger protein (MAZ), jun-D proto-oncogene (JUND), and ELF1. These transcription factor binding sites tended to be given higher predictive importance at lower resolutions. In fact, at lower resolutions, the predictive importance was found to be much more evenly dispersed across the top transcription factor binding sites, than for higher resolutions. This is again suggestive of the fact that the spatial relationship between functional annotations, specifically transcription factor binding sites, and TAD boundaries changes at different Hi-C data resolutions.
